\showyourwork: a workflow for open source scientific articles

### 2022-10-21 src/tex/ms.tex
todos

% Imports
\usepackage{showyourwork}
\usepackage{etoolbox}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage[figuresright]{rotating}
\usepackage{ifthen}

@@

% Dev & review
\newcommand\xxx[1]{{\color{red}#1}}
\newcommand\needscite{{\color{red}(citation needed)}}

@@

%
\section{Three arguments for reproducibility}
\label{sec:args}
\xxx{Short intro here. I can think of three big arguments for reproducibility, but there are likely many more.}

% 1/3
\subsection{Ensuring the validity of results}
\label{sec:args:validity}
As astronomical research software becomes increasingly more complex, and as research results become increasingly more interdependent, it becomes ever more challenging to ensure the validity and correctness of results published in the literature. 
Unfortunately, the current peer review system in astronomy is simply not set up to do this.
Checking all of the results in a paper would require the painstaking and methodical review of all of the paper's methods---which usually means scrutinizing all of the code used to generate the figures, tables, and other quantities in the paper. 
%
\xxx{Any stats on how much time a typical referee spends on reviewing a paper? Would be nice to back all this up with some numbers.}
%
\xxx{We could also look for papers with erroneous results that propagated outwards in the literature because some big mistake made it past peer review. Anyone know of any classic examples?}
%
In practice, this is virtually impossible for three reasons:
%
\begin{enumerate}
    %
    \item Modern codebases can be very large and often require deep familiarity with the software to use---not to mention review them. Volunteer referees rarely have the time to invest in learning new software in order to provide a comprehensive review.
    %
    \item  Writing a paper in astronomy is rarely ever done in a linear, procedural fashion: the codebase is constantly changing, and the state of the code when (say) Figure 1 was produced may be very different from that when (say) Figure 2 was made. 
    Moreover, many results depend on the execution of lengthy pipelines with intermediate steps, each potentially requiring manual tinkering that is not always documented and may be difficult to replicate exactly.
    %
    \item The majority of astronomical code is not open source and simply cannot be vetted by third parties \needscite. 
    While there has been a marked increase in the number of open source astronomical tools in recent years (e.g., \astropy, \exoplanet, \emcee, \exofast...), most code associated with the generation of the results in individual papers is not open source \needscite; readers are often expected to take it on faith that there are no bugs in that code, or that the code works exactly as described in the text, with no pitfalls or missing details. 
    Even when the code is made publicly available, e.g., by being published on \GitHub, it is often not documented sufficiently to enable one to execute it and reproduce the paper's results out-of-the-box. 
    And even with proper documentation, the code may require external dependencies, custom virtual environments, or access to closed-source datasets that make it difficult or impossible for a third party to replicate it.
    %
\end{enumerate}
%
\xxx{A few sentences here on how an easily reproducible article could help with all this.}

% 2/3
\subsection{Reducing duplication of effort}
\label{sec:args:duplication}
\xxx{We shouldn't have to ever reinvent the wheel. Astronomy should be collaborative. Time saved on duplicated effort can be invested in moving the field forward rather than always playing catch-up.}
%
\xxx{Perhaps some stats on CPU usage in the field, and how reproducibility can help?}
%
\xxx{Can also talk about open data here; e.g., telescope time is super competitive; we should make it easy for others to use our data, both raw and processed.}

% 3/3
\subsection{Promoting equity and inclusivity}
\label{sec:args:dei}
\xxx{Proprietary legacy code is antithetical to DEI. ``Source code available upon request'' is antithetical to DEI. ``My code is on my website'' is antithetical to DEI. It's 2022, FFS. We need to do everything we can to make astronomy open to everyone.}

\section{Current approaches to reproducibility}
\label{sec:intro:curr}
\xxx{Discuss code links in \citet{Luger2019,Luger2021a,Luger2021b}, and how they have caught on, e.g., \citet{Paillas2022}.} 
%
\xxx{Give a detailed overview of reproducibility in astronomy; \href{https://maneage.org/pdf/slides-intro.pdf}{this page} has some nice resources. Look at reviews by Harlan Krumholz in medicine. Talk about \maneage.}
%

@@

\subsection{Repository structure}
\label{sec:usage:struct}
%
\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=0.8\linewidth]{figures/tree.pdf}
        \caption{
            The basic repository structure for an open source scientific article built with \syw.
            The article you are reading was generated from a repository with this exact structure; you can check it out \href{\commiturl}{here}.
            This figure was automatically generated from the \TikZ code in \repoargfile{src/scripts/tree.tex} by specifying a custom command in \repoargfile{showyourwork.yml}.
        }
        \label{fig:tree}
        \script{tree.tex}
    \end{centering}
\end{figure}
%
Figure~\ref{fig:tree} shows the basic directory structure for a \syw article repository as of version \VERSION{}.
Repositories are comprised of configuration files at the root level and a directory called \repoargfile{src} containing the manuscript files and all of the the code and workflow scripts needed to render the final article \pdf file.
%
\xxx{More details here.}

\section{How it works}
\label{sec:how-it-works}
\xxx{Discuss how things are actually implemented. Talk about the dependence on \snakemake \citep{Molder2021} and \tectonic. Discuss the integration with \Zenodo for both static datasets and cached datasets. Discuss integration with \Overleaf.}

@@

Figure~\ref{fig:two_moons} is an example of a figure depending on an intermediate result that is cached on Zenodo Sandbox.
\xxx{More details.}

\xxx{Discuss Table~\ref{tab:variable} and variable output.}
Variable output can also occur inline. For instance, the following sentence is programmatically generated from the same script: 
\variable{output/kernel_value.tex} 
Note the icon in the margin that automatically links to the corresponding \snakemake rule, where users can easily change the values of the kernel parameters.

\xxx{Discuss Figure~\ref{fig:dag} and this paper as a whole as a meta-example.}

\xxx{Finally, cite a bunch of papers here that already use showyourwork and showcase its various features.}

\section{Current issues and future work}
\label{sec:future}
\xxx{This is a work in progress. Currently focused primarily on short-term reproducibility; discuss the various issues that arise from relying on \GitHub, \conda, \pip, etc. Broken links will be an issue. Conda environments are not impermeable. Currently difficult (impossible?) to run stuff on a cluster. Doesn't play nice with journals. Anything  else?}

\section{Conclusion}
\label{sec:conclusion}
\xxx{Stuff here. showyourwork is a community project -- please contribute!}

@@

\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=0.75\linewidth]{figures/two_moons.pdf}
        \caption{
            A normalizing flow demonstrated on the two moons data set from \scikitlearn;
            reproduced from Figure 1 in \citet{Crenshaw2022}.
            This figure was automatically generated from the script \repoargfile{src/scripts/two\_moons.py}
            and an intermediate dataset that was automatically cached on \Zenodo.
            \xxx{@jfcrenshaw: is it ok for us to reproduce this figure of yours?}
        }
        \label{fig:two_moons}
        \script{two_moons.py}
    \end{centering}
\end{figure}

\begin{table}
    \begin{center}
        \begin{tabular}{@{}l|l@{}}
            \hline
            Name & Representation \\
            \hline
            \variable{output/kernels.tex}
        \end{tabular}
    \end{center}
    \caption{The functional forms of various commonly used Gaussian process kernels; reproduced from Table~1 in \citet{Aigrain2022}.
    The contents of this table are programmatically generated; the kernels are defined as \sympy expressions and rendered into \TeX form in the script \repoargfile{src/scripts/kernels.py}.
    The autogenerated \GitHub icon in the margin links to the rule in the \repoargfile{Snakefile} that runs that script.
    \xxx{@dfm: is it ok for us to reproduce this table of yours?}}
    \label{tab:variable}
\end{table}

\begin{sidewaysfigure}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/dag.pdf}
        \caption{
            A directed acyclic graph (DAG) showing the complete list of dependencies for the article. 
            Scripts are shown in green, \Zenodo deposits in blue, datasets in purple, and \TeX files in red.
            This figure is located in the \repoargfile{src/static} directory and, unlike the other figures in this article, is version controlled by \git. The \repoargfile{src/static} directory is reserved for figures that are not programmatically generated and simply get copied over to the output directory at compile time.
            \xxx{TODO: Regenerate me!}
        }
        \label{fig:dag}
    \end{centering}
\end{sidewaysfigure}

>>>>

% Imports
\usepackage{showyourwork}
\usepackage{etoolbox}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage[figuresright]{rotating}
\usepackage{ifthen}
\usepackage{xparse}
\usepackage{stackengine}

@@

% Dev & review
\newcommand\needscite{{\color{red}(citation needed)}}
\DeclareDocumentCommand{\todo}{o o o o o o o o m}{%
    {\color{red}#9}%
    \marginpar{%
        \raisebox{-0.5em}{%
            \setstackgap{S}{0pt}%
            \Shortstack[c]{
                \IfValueT{#1}{\includegraphics[width=1.75em]{avatars/#1.png}\IfValueT{#2}{\includegraphics[width=1.75em]{avatars/#2.png}}}
                \IfValueT{#3}{\includegraphics[width=1.75em]{avatars/#3.png}\IfValueT{#4}{\includegraphics[width=1.75em]{avatars/#4.png}}}
                \IfValueT{#5}{\includegraphics[width=1.75em]{avatars/#5.png}\IfValueT{#6}{\includegraphics[width=1.75em]{avatars/#6.png}}}
                \IfValueT{#7}{\includegraphics[width=1.75em]{avatars/#7.png}\IfValueT{#8}{\includegraphics[width=1.75em]{avatars/#8.png}}}
            }%
        }
    }%
}

@@

%
\section{Three arguments for reproducibility}
\label{sec:args}
\todo[rodluger][adrn][dfm][milescranmer][jfcrenshaw][mathren][katiebreivik][tomwagg]{Short intro here. I can think of three big arguments for reproducibility, but there are likely many more.}

% 1/3
\subsection{Ensuring the validity of results}
\label{sec:args:validity}
As astronomical research software becomes increasingly more complex, and as research results become increasingly more interdependent, it becomes ever more challenging to ensure the validity and correctness of results published in the literature. 
Unfortunately, the current peer review system in astronomy is simply not set up to do this.
Checking all of the results in a paper would require the painstaking and methodical review of all of the paper's methods---which usually means scrutinizing all of the code used to generate the figures, tables, and other quantities in the paper. 
%
\todo[adrn][dfm][milescranmer][mathren]{
    Any stats on how much time a typical referee spends on reviewing a paper? Would be nice to back all this up with some numbers.
    We could also look for papers with erroneous results that propagated outwards in the literature because some big mistake made it past peer review. Anyone know of any classic examples?
}

%
In practice, this is virtually impossible for three reasons:
%
\begin{enumerate}
    %
    \item Modern codebases can be very large and often require deep familiarity with the software to use---not to mention review them. Volunteer referees rarely have the time to invest in learning new software in order to provide a comprehensive review.
    %
    \item  Writing a paper in astronomy is rarely ever done in a linear, procedural fashion: the codebase is constantly changing, and the state of the code when (say) Figure 1 was produced may be very different from that when (say) Figure 2 was made. 
    Moreover, many results depend on the execution of lengthy pipelines with intermediate steps, each potentially requiring manual tinkering that is not always documented and may be difficult to replicate exactly.
    %
    \item The majority of astronomical code is not open source and simply cannot be vetted by third parties \needscite. 
    While there has been a marked increase in the number of open source astronomical tools in recent years (e.g., \astropy, \exoplanet, \emcee, \exofast...), most code associated with the generation of the results in individual papers is not open source \needscite; readers are often expected to take it on faith that there are no bugs in that code, or that the code works exactly as described in the text, with no pitfalls or missing details. 
    Even when the code is made publicly available, e.g., by being published on \GitHub, it is often not documented sufficiently to enable one to execute it and reproduce the paper's results out-of-the-box. 
    And even with proper documentation, the code may require external dependencies, custom virtual environments, or access to closed-source datasets that make it difficult or impossible for a third party to replicate it.
    %
\end{enumerate}
%
\todo[rodluger]{A few sentences here on how an easily reproducible article could help with all this.}

% 2/3
\subsection{Reducing duplication of effort}
\label{sec:args:duplication}
\todo[rodluger]{We shouldn't have to ever reinvent the wheel. Astronomy should be collaborative. Time saved on duplicated effort can be invested in moving the field forward rather than always playing catch-up.}
%
\todo[adrn][dfm]{Perhaps some stats on CPU usage in the field, and how reproducibility can help?}
%
\todo[adrn][dfm]{Can also talk about open data here; e.g., telescope time is super competitive; we should make it easy for others to use our data, both raw and processed.}

% 3/3
\subsection{Promoting equity and inclusivity}
\label{sec:args:dei}
\todo[rodluger][adrn]{Proprietary legacy code is antithetical to DEI. ``Source code available upon request'' is antithetical to DEI. ``My code is on my website'' is antithetical to DEI. It's 2022, FFS. We need to do everything we can to make astronomy open to everyone.}

\section{Current approaches to reproducibility}
\label{sec:intro:curr}
\todo[rodluger]{Discuss code links in \citet{Luger2019,Luger2021a,Luger2021b}, and how they have caught on, e.g., \citet{Paillas2022}.} 
%
\todo[rodluger][adrn][dfm]{Give a detailed overview of reproducibility in astronomy; \href{https://maneage.org/pdf/slides-intro.pdf}{this page} has some nice resources. Look at reviews by Harlan Krumholz in medicine. Talk about \maneage.}
%

@@

\subsection{Repository structure}
\label{sec:usage:struct}
%
\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=0.5\linewidth]{figures/tree.pdf}
        \caption{
            The basic repository structure for an open source scientific article built with \syw.
            The article you are reading was generated from a repository with this exact structure; you can check it out \href{\commiturl}{here}.
            This figure was automatically generated from the \TikZ code in \repoargfile{src/scripts/tree.tex} by specifying a custom command in \repoargfile{showyourwork.yml}.
        }
        \label{fig:tree}
        \script{tree.tex}
    \end{centering}
\end{figure}
%
Figure~\ref{fig:tree} shows the basic directory structure for a \syw article repository as of version \VERSION{}.
Repositories are comprised of configuration files at the root level and a directory called \repoargfile{src} containing the manuscript files and all of the the code and workflow scripts needed to render the final article \pdf file.
%
\todo[rodluger]{More details here.}

\section{How it works}
\label{sec:how-it-works}
\todo[rodluger]{Discuss how things are actually implemented. Talk about the dependence on \snakemake \citep{Molder2021} and \tectonic. Discuss the integration with \Zenodo for both static datasets and cached datasets. Discuss integration with \Overleaf.}

@@

Figure~\ref{fig:two_moons} is an example of a figure depending on an intermediate result that is cached on Zenodo Sandbox.
\todo[rodluger]{More details.}

\todo[rodluger]{Discuss Table~\ref{tab:variable} and variable output.}
Variable output can also occur inline. For instance, the following sentence is programmatically generated from the same script: 
\variable{output/kernel_value.tex}
Note the icon in the margin that automatically links to the corresponding \snakemake rule, where users can easily change the values of the kernel parameters.

\todo[rodluger]{Discuss Figure~\ref{fig:dag} and this paper as a whole as a meta-example.}

\todo[rodluger]{Finally, cite a bunch of papers here that already use showyourwork and showcase its various features.}

\section{Current issues and future work}
\label{sec:future}
\todo[rodluger][dfm][mathren][milescranmer]{This is a work in progress. Currently focused primarily on short-term reproducibility; discuss the various issues that arise from relying on \GitHub, \conda, \pip, etc. Broken links will be an issue. Conda environments are not impermeable. Currently difficult (impossible?) to run stuff on a cluster. Doesn't play nice with journals. Anything  else?}

\section{Conclusion}
\label{sec:conclusion}
\todo[rodluger][dfm][adrn][milescranmer][jfcrenshaw][mathren][katiebreivik][tomwagg]{Stuff here. showyourwork is a community project -- please contribute!}

@@

\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=0.75\linewidth]{figures/two_moons.pdf}
        \caption{
            A normalizing flow demonstrated on the two moons data set from \scikitlearn;
            reproduced from Figure 1 in \citet{Crenshaw2022}.
            This figure was automatically generated from the script \repoargfile{src/scripts/two\_moons.py}
            and an intermediate dataset that was automatically cached on \Zenodo.
        }
        \label{fig:two_moons}
        \script{two_moons.py}
    \end{centering}
\end{figure}

\begin{table}
    \begin{center}
        \begin{tabular}{@{}l|l@{}}
            \hline
            Name & Representation \\
            \hline
            \variable{output/kernels.tex}
        \end{tabular}
    \end{center}
        \caption{The functional forms of various commonly used Gaussian process kernels; based on Table~1 in \citet{Aigrain2022}.
        The contents of this table are programmatically generated; the kernels are defined as \sympy expressions and rendered into \TeX form in the script \repoargfile{src/scripts/kernels.py}.
        The autogenerated \GitHub icon in the margin links to the rule in the \repoargfile{Snakefile} that runs that script.
    }
    \label{tab:variable}
\end{table}

\begin{sidewaysfigure}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/dag.pdf}
        \caption{
            A directed acyclic graph (DAG) showing the complete list of dependencies for the article. 
            Scripts are shown in green, \Zenodo deposits in blue, datasets in purple, and \TeX files in red.
            This figure is located in the \repoargfile{src/static} directory and, unlike the other figures in this article, is version controlled by \git. The \repoargfile{src/static} directory is reserved for figures that are not programmatically generated and simply get copied over to the output directory at compile time.
        }
        \label{fig:dag}
    \end{centering}
\end{sidewaysfigure}
